Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland

### 2017-09-14 paper/main.tex
WTK final comments

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 15 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that few clusters could be fully characterized by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain the homeostasis of adult mammary epithelial cells.}
\newpage

@@

We isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI).
For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (Supplementary Figure 1, \autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility between the biological replicates (Supplementary Figure 2,3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d, Supplementary Table 1).\\

@@

In the luminal compartment we found two larger subgroups (\autoref{F2}a).
C1-C5 showed characteristics of hormone-sensing cells (Hs) such as high expression levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c).
Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g. \textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).
In contrast, C8-C11 expressed expressed milk proteins (e.g. \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c and \autoref{T1}).
Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c).

In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}b-c).
In addition, it also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).

@@

The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of cells from C6 (Lp).  
We further noted that the left arm of the differentiation trajectory terminates at C8 (Avd) and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
In between C6 and C8  cells mainly derived from C10 (Avp, \autoref{F3}b) .
On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Esr1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).

@@

We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (see materials and methods for definition, Supplementary Table 2).
These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
\textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
In addition, we identified 1005 genes with branch-specific expression patterns.
We then clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f, Supplementary Table 3, Supplementary Figure 6).
On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}, \autoref{F4}f) .
The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g. \textit{Olah}, \textit{Cd36}, \autoref{F4}f).
Of note, milk proteins vary in their time of activation, with caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were expressed in the final stages.
Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g. \textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g. \textit{Vdr} and \textit{Areg}).
During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
Amongst the branch-specific transcription factors we found, for example, \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F4}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F4}d).
\textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F4}d).

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage.
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterized by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3} and \autoref{F4}.
This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that cells from C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing or secretory cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells.
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and lactation.
We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 5--------------------------------------------------------------
\textbf{Figure 5. The effect of parity on the transcriptomic landscape of the luminal progenitor compartment}
\fakefigure{F5}
\textbf{(a)} Comparison of fold changes from C6 versus luminal compartment and fold changes from C7 versus the luminal cells.
The genes represent the top 500 differentially expressed genes between C6 and luminal cells.
\textbf{(b)} Volcano plot illustrating differential expression between C6 and C7, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
P values are adjusted for multiple testing using Benjamini-Hochberg.
\textbf{(c-d)} Visualization of expression difference for genes related to lactation \textbf{(c)} or immunity \textbf{(d)} for the six luminal clusters in NP and PI. Expression values correspond to normalised UMI counts.
\textbf{(e)} Cells from PI projected on the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
%-------------------------------------------------------------------------------------------------------------------

@@

\textbf{Supplementary Figure 2. Quality control of sequencing data}
\textbf{(a)} Table summarizing quality control criteria per sample.
Number of unique molecules, genes detected and number of reads represent the median value for each sample.
\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)}.
\textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules.
For choice of thresholds see materials and methods.

@@

\textbf{Supplementary Figure 7. C7 is a post-parity luminal progenitor cluster}
\textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd, Hsp and Lp.
\textbf{(b)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
The PCA was computed on the top 500 differentially expressed genes between Lp and the rest of the luminal NP gland.
PC1 separates the progenitors and the differentiated cells, with the progenitors showing negative PC1 values.
\textbf{(c)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
\textbf{(d)} Top GO-terms (biological processes) that are significantly enriched in upregulated genes from \textbf{(a)}.
Dashed line indicates P value threshold at 0.001.

@@

\textbf{Table S4.} Genes differentially expressed between C6 and C7

@@

In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
Libraries prepared from all time-points showed high quality that was reproducible between the two biological replicates (Supplementary Figure 2,3).
We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each time-point, while requiring a minimum value of 1,000 total molecules and 500 genes detected.  
This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Figure 2d).
We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017, Griffiths2017}.
For this we excluded cells with barcodes that appeared in more than one sample (non-unique barcodes).
The reasoning behind this being that any index swapped read between samples will also share the same cellular barcode.
Exclusion of these cells hence offers protection against index swapping.
This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
Gene expression values were then normalised by size factors, that were estimated as previously described\autocite{Lun2016}.
The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and a Benjamini-Hochberg adjusted P value smaller than or equal to 0.05 were defined as HVG.
The t-SNE embeddings were computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).

@@

A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize various genesets in the analysis.
The genes of interest were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.

@@

All computational analyses were performed in R (Version 3.4.1) using standard functions unless otherwise indicated.
Code is available online at \url{https://github.com/MarioniLab/MammaryGland}.
Data can be explored at \url{https://karstenbach.shinyapps.io/webapp/}.

>>>>

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 15 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that few clusters could be fully characterised by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain the homeostasis of adult mammary epithelial cells.}
\newpage

@@

We isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI).
For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (Supplementary Figure 1, \autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility between the biological replicates (Supplementary Figure 2,3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach (see materials and methods).
This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d, Supplementary Table 1).\\

@@

In the luminal compartment we found two larger subgroups (\autoref{F2}a).
C1-C5 showed characteristics of hormone-sensing cells (Hs) such as high expression levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c).
Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g. \textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting restricted progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).
In contrast, C8-C11 expressed proteins (e.g. \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c and \autoref{T1}).
Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a restricted alveolar progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated alveolar state (Avd) (\autoref{F2}b-c).

In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}b-c).
In addition, the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).

@@

We found the expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of cells from C6 (Lp).  
We further noted that the left arm of the differentiation trajectory terminates at C8 (Avd) and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
In between C6 and C8, we found cells that were mainly derived from C10 (Avp, \autoref{F3}b).
On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Esr1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).

@@

We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (see materials and methods for definition, Supplementary Table 2).
These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8}\autocite{Fu2017} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
\textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
In addition, we identified 1005 genes with branch-specific expression patterns.
We then clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f, Supplementary Table 3, Supplementary Figure 6).
On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}, \autoref{F4}f).
The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g. \textit{Olah}, \textit{Cd36}, \autoref{F4}f).
Of note, milk genes vary in their time of activation, with caseins switched on earler during differentiation, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were only expressed in the final stages.
Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g. \textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g. \textit{Vdr} and \textit{Areg}).
During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
Amongst the branch-specific transcription factors we found, for example, \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F4}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F4}d).
\textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
The expression of the known progenitor master regulator \textit{Elf5}\autocite{Oakes2008,Rios2014} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F4}d).

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage.
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterised by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3} and \autoref{F4}.
This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
In this study we could not detect any cells in transition between the basal and luminal cells.
Therefore, we have no evidence to support the contribution of a putative multipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that cells from C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing or secretory cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells.
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and lactation.
We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 5--------------------------------------------------------------
\textbf{Figure 5. The effect of parity on the transcriptomic landscape of the luminal progenitor compartment}
\fakefigure{F5}
\textbf{(a)} Comparison of fold changes from C6 versus luminal compartment and fold changes from C7 versus the luminal cells.
The genes represent the top 500 differentially expressed genes between C6 and luminal cells.
\textbf{(b)} Volcano plot illustrating differential expression between C6 and C7, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
P values are adjusted for multiple testing using Benjamini-Hochberg.
\textbf{(c-d)} Visualization of expression difference for genes related to lactation \textbf{(c)} or immunity \textbf{(d)} for the six luminal clusters in NP and PI. Expression values correspond to normalised UMI counts.
\textbf{(e)} Cells from PI projected on the diffusion map from the NP and G time-point. Cells from NP and G are coloured in grey.
%-------------------------------------------------------------------------------------------------------------------

@@

\textbf{Supplementary Figure 2. Quality control of sequencing data}
\textbf{(a)} Table summarising quality control criteria per sample.
Number of unique molecules, genes detected and number of reads represent the median value for each sample.
\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)}.
\textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules.
For choice of thresholds see materials and methods.

@@

\textbf{Supplementary Figure 7. C7 is a post-parity luminal progenitor cluster}
\textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd (C1 vs C2), Hsp (C3 vs C4) and Lp (C7 vs. C6).
\textbf{(b)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
The PCA was computed on the top 500 differentially expressed genes between Lp and the rest of the luminal NP gland.
PC1 separates the progenitors and the differentiated cells, with the progenitors showing negative PC1 values.
\textbf{(c)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
\textbf{(d)} Top GO-terms (biological processes) that are significantly enriched in upregulated genes from the same differential expression in analysis visualized in \textbf{(a)}.
Dashed line indicates P value threshold at 0.001.

@@

@@

In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
Libraries prepared from all time-points showed high quality that was reproducible between the two biological replicates (Supplementary Figure 2,3).
We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each time-point, while requiring a minimum value of 1,000 total molecules and 500 genes detected.  
This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules: 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Figure 2d).
We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017, Griffiths2017}.
For this we excluded cells with barcodes that appeared in more than one sample (non-unique barcodes).
The reasoning behind this being that any index swapped read between samples will also share the same cellular barcode.
Exclusion of these cells hence offers protection against index swapping.
This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
Gene expression values were then normalised by size factors, that were estimated as previously described\autocite{Lun2016}.
The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and a Benjamini-Hochberg adjusted P value smaller than or equal to 0.05 were defined as HVG.
The t-SNE embeddings were computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).

@@

A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterise various genesets in the analysis.
The genes of interest were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.

@@

All computational analyses were performed in R (Version 3.4.1) using standard functions unless otherwise indicated.
Code is available online at \url{https://github.com/MarioniLab/MammaryGland}.
Data can be explored at \url{http://marionilab.cruk.cam.ac.uk/mammaryGland}.
